### 2016-09-29 paper2/paper.tex
paper 2 template

\documentclass[letterpaper, twocolumn,11pt]{article}
\usepackage{epsfig,endnotes,amsmath,usenix}

\title{\textbf{Regressions And Stuff}}

\author{
\rm{Vahid Fazel-Rezai}\\
\rm{\texttt{vfazel@mit.edu}}
\and 
\rm{Akshay Ravikumar}\\
\rm{\texttt{akshayr@mit.edu}}
}

\date{May 6, 2016}

\begin{document}
% \begin{titlepage}
% \centering
% Matei Zaharia\\
% TR 10\\
% May 6, 2016
% \end{titlepage}

\maketitle
\section{Introduction}

In this paper, we discuss various methods of calculating regression. First, we discuss the benfits of using Stochastic Gradient Descent versus Batch Gradient Descent. Next, we analyze the performance of these methods in performing a Least Squares Linear regression. Finally, we discuss blah blah

\section{Overview}

\section{Gradient Descent}
\subsection{Batch Gradient Descent}
\subsection{Stochastic Gradient Descent}
\subsection{Effect of Finite Gradient}

\end{document}

### 2016-09-29 code/P1/plot.py
more problem 1 stuff

@@ -1,0 +1,18 @@
+import matplotlib.pyplot as plt
+
+x = [(1000.0, 1e-10, 1.407205159025751), (1000.0, 1.18e-10, 1.5293301960470953), (1000.0, 1.36e-10, 1.6423311195602779), (1000.0, 1.54e-10, 1.7483084457705826), (1000.0, 1.7200000000000003e-10, 1.8481359485031223), (1000.0, 1.9000000000000002e-10, 1.943118354878313), (1000.0, 2.08e-10, 2.0338965577357535), (1000.0, 2.2600000000000005e-10, 2.1207986289140495), (1000.0, 2.4400000000000004e-10, 2.2043846443844473), (1000.0, 2.62e-10, 2.285072657125581), (1000.0, 2.8e-10, 2.363439188966732), (1000.0, 2.9800000000000006e-10, 2.439053512614124), (1000.0, 3.1600000000000005e-10, 2.5122845198739605), (1000.0, 3.3400000000000004e-10, 2.5840060973242878), (1000.0, 3.520000000000001e-10, 2.6535498635515475), (1000.0, 3.7000000000000007e-10, 2.7214962472571194), (1000.0, 3.8800000000000006e-10, 2.7880720090201456), (1000.0, 4.0600000000000005e-10, 2.827534051606957), (1000.0, 4.2400000000000004e-10, 2.827534051606957), (1000.0, 4.420000000000001e-10, 2.827534051606957), (1000.0, 4.6000000000000007e-10, 2.827534051606957), (1000.0, 4.78e-10, 2.827534051606957), (1000.0, 4.96e-10, 2.827534051606957), (1000.0, 5.140000000000001e-10, 2.827534051606957), (1000.0, 5.32e-10, 2.827534051606957), (1000.0, 5.500000000000001e-10, 2.827534051606957), (1000.0, 5.68e-10, 2.827534051606957), (1000.0, 5.86e-10, 2.827534051606957), (1000.0, 6.040000000000001e-10, 2.827534051606957), (1000.0, 6.22e-10, 2.827534051606957), (1000.0, 6.400000000000001e-10, 2.827534051606957), (1000.0, 6.58e-10, 2.827534051606957), (1000.0, 6.76e-10, 2.827534051606957), (1000.0, 6.940000000000001e-10, 2.827534051606957), (1000.0, 7.12e-10, 2.827534051606957), (1000.0, 7.300000000000001e-10, 2.827534051606957), (1000.0, 7.
[…]

### 2016-09-29 code/P1/problem1.py

@@ -2,8 +2,9 @@ import pylab as pl
 import numpy as np
 import math
 import matplotlib.pyplot as plt
+import time
 
-def getData():
+def getGradientData():
 
     # load the parameters for the negative Gaussian function and quadratic bowl function
     # return a tuple that contains parameters for Gaussian mean, Gaussian covariance,
@@ -19,18 +20,47 @@ def getData():
 
     return (gaussMean, gaussCov, quadBowlA, quadBowlb) 
 
+def getFittingData():
+    
+    # load the fitting data for X and y and return as elements of a tuple
+    # X is a 100 by 10 matrix and y is a vector of length 100
+    # Each corresponding row for X and y represents a single data sample
 
-def gradient_descent(objective_func, gradient_func, guess, step, threshold, plot = False):
+    X = pl.loadtxt('fittingdatap1_x.txt')
+    y = pl.loadtxt('fittingdatap1_y.txt')
+
+    return (X,y) 
+
+
+def gradient_descent(objective_func, gradient_func, guess, step, threshold):
 	prev = None
 	delta = float("inf")
 	mags = []
 	while delta > threshold:
+	# for i in range(2000):
 		delta = abs(objective_func(guess) - objective_func(prev)) if prev is not None else float("inf")
 		gradient = gradient_func(guess)
 		mags.append(np.linalg.norm(gradient))
 		prev = guess
 		guess = np.subtract(guess, step * gradient)
 	return guess, mags
+
+def stochastic_gradient_descent(objective_func, gradient_i, n, guess, threshold):
+	prev = None
+	delta = float("inf")
+	mags = []
+	r = 1
+	n_t = lambda r : (r*1000 + 10000)**(-0.9)
+	while delta > threshold:
+		delta = abs(objective_func(guess) - objective_func(prev)) if prev is not None else float("inf")
+		prev = guess
+		for i in np.random.permutation(n):
+			rate = n_t(r)
+			gradient = gradient_i(guess, i)
+			guess = np.subtract(guess, rate * gradient)
+			r += 1
+	return guess, mags
+
 
 def gaussian(mean, cov, delta):
 	def gaussian(x):
[…]

### 2016-09-29 paper2/paper.tex
intro to gradient descent

@@
-\documentclass[letterpaper, twocolumn,11pt]{article}
+\documentclass[letterpaper, twocolumn,10pt]{article}
 \usepackage{epsfig,endnotes,amsmath,usenix}
@@
 
-\author{
-\rm{Vahid Fazel-Rezai}\\
-\rm{\texttt{vfazel@mit.edu}}
-\and 
-\rm{Akshay Ravikumar}\\
-\rm{\texttt{akshayr@mit.edu}}
-}
+\author{}
 
@@
 \section{Gradient Descent}
+
+\emph{Gradient descent} is an iterative optimization algorithm; in other words, it calculates the parameters $\mathbf{x}$ that minimize a given objective function $f(\mathbf{x})$.
+The algorithm repeatedly translates an initial \emph{guess} $\mathbf{x}_0$ in a direction proportional to the negative \emph{gradient} $\nabla f(x)$.
+
+In every step of the itreration, we update our guess as following:
+\begin{align*}
+\mathbf{x}_{n+1} = \mathbf{x} - \lambda \nabla f(\mathbf{x}_{n})
+\end{align*}
+where $\lambda$ is the \emph{step size} of the iteration. The algorith terminates upon the following convergence condition: 
+\begin{align*}
+|f(\mathbf{x}_{n+1}) - f(\mathbf{x}_{n})| < \delta
+\end{align*}
+where $\delta$ is the convergence \emph{threshold}. Upon convergence, the algorithm returns a final guess of $\mathbf{x}_{\text{opt}} = \mathbf{x}_{n+1}$.
+
+In this section, we compare two methods of gradient descent: \emph{batch gradient descent}, which computes a gradient over all samples for each iteration, and \emph{stochastic gradient descent}, which uses pointwise gradients instead.
+
+To analyze the perforamnce of the different algorithms, we use the \emph{Gaussian function}:
+\begin{align*}
+f(x) = - \dfrac{1}{\sqrt{(2\pi)^n |\sigma|}} \exp \left[-\dfrac{1}{2} (x - u)^T \sigma^{-1} (x-u)\right]
+\end{align*}
+In addition, we use the \emph{quadratic bowl function}:
+\begin{align*}
+f(x) = \dfrac{1}{2} x^T A x - x^T b 
+\end{align*}
+Finally, we use the \emph{least squares error} function:
+\begin{align*}
+J(\theta) = |X\theta - y|^2
+\end{align*}
+
+We analyze the affects of step size ($\lambda$), threshold ($\delta$), and initial guess $\mathbf{x}_{0}$ on the performance and accuracy of the gradient descent algorithm.
+
 \subsection{Batch Gradient Descent}
+
 \subsection{Stochastic Gradient Descent}

### 2016-09-30 paper2/paper.tex
more problem 1 stuff

@@
-\documentclass[letterpaper, twocolumn,10pt]{article}
-\usepackage{epsfig,endnotes,amsmath,usenix}
+\documentclass{paper}
+\usepackage{fullpage, amsmath, graphicx}
 
@@
 \end{align*}
-where $\lambda$ is the \emph{step size} of the iteration. The algorith terminates upon the following convergence condition: 
+where $\lambda$ is the \emph{step size} of the iteration. The algorithm terminates upon the following convergence condition: 
 \begin{align*}
@@
 
-In this section, we compare two methods of gradient descent: \emph{batch gradient descent}, which computes a gradient over all samples for each iteration, and \emph{stochastic gradient descent}, which uses pointwise gradients instead.
+In this section, we compare two methods of gradient descent: \emph{batch gradient descent}, which computes a gradient over all samples for each iteration, and \emph{stochastic gradient descent}, which instead uses pointwise gradients.
 
@@
 \begin{align*}
-f(x) = - \dfrac{1}{\sqrt{(2\pi)^n |\sigma|}} \exp \left[-\dfrac{1}{2} (x - u)^T \sigma^{-1} (x-u)\right]
+f(x) = - \dfrac{1}{\sqrt{(2\pi)^n |\Sigma|}} \exp \left[-\dfrac{1}{2} (x - u)^T \Sigma^{-1} (x-u)\right]
 \end{align*}
@@
 
-We analyze the affects of step size ($\lambda$), threshold ($\delta$), and initial guess $\mathbf{x}_{0}$ on the performance and accuracy of the gradient descent algorithm.
-
 \subsection{Batch Gradient Descent}
 
+In batch gradient descent, for every iteration we calculate the gradient across the entire sample set.
+In this section, we analyze the affects of step size ($\lambda$), threshold ($\delta$), and initial guess $\mathbf{x}_{0}$ on the performance and accuracy of the gradient descent algorithm.
+
+First, we note that step size has a significant impact on the convergence rate of the algorithm.
+We run gradient descent on an arbitrary quadratic bowl function.
+As we increase the value of $\lambda$ from $0.0001$ to $0.001$, the magnitude of the gradient decreases more quickly, and the algorithm converges faster.
+Note that in this figure, all of the gradient descents were run with a fixed size of 2000 iterations to illustrate the effects on the gradient over time.
+
+\begin{figure}[ht!]
+  \centering
+    \includegraphics[width=0.5\textwidth]{../images/quadratic_steps_vs_mag}
+  \caption{As the step size increases, the algorithm converges quicker.}
+\end{figure}
+
+In addition, the step size and threshold control the accuracy of the gradient descent.
+As step size decreases, we expect a more accurate result because the function of the initial guess over time is smoother.
+Similarly, as the threshold decreases, we expect a more accurate result because the algorithm needs to run for longer before reaching a stable equilibrium.
+To verify these claims, we run gradient descent using a two-dimensional Gaussian with mean $(10, 10)$.
+As we vary $\lambda$ and $\delta$, we plot the least-squares error between the returned value and the actual optimum of $(10, 10)$ (see Figure 2).
+Indeed, the error decreases as $\lambda$ and $\delta$ decrease.
+
+\begin{figure}[ht!]
+  \centering
+    \includegraphics[width=0.5\textwidth]{../images/lsq_error_gradient}
+  \caption{As $\lambda$ and $\delta$ decrease, error decreases.}
+\end{figure}
+
+Finally, we observe the affect of the initial case on the rate of convergence.
+When running gradient descent on a two-dimensional Gaussian with mean $(10, 10)$, one would expect that as the intial guess appraoches the optimal value, the convergence rate increases.
+Indeed, running the algorithm with $\lambda = 1000$ and $\delta = 10^{-11}$, we notice that the rate of convergence is much higher as the initial guess approaches the actual value (see Figure 3).
+
+\begin{figure}[ht!]
+  \centering
+    \includegraphics[width=0.5\textwidth]{../images/guess_vs_mags_2}
+  \caption{As $\mathbf{x}_0$ appraoches $(10, 10)$, the rate of convergence increases.}
+\end{figure}
+
 \subsection{Stochastic Gradient Descent}
+
 \subsection{Effect of Finite Gradient}
 
+When the objective function doesn't have a closed form solution, one can approximate the gradient using a central finite difference. In other words, we have 
+\begin{align*}
+\nabla f(x) \approx \nabla_h[f](x) \dfrac{f(x+h/2) - f(x-h/2)}{h}
+\end{align*}
+Therefore, we only need the objective function to approximate the gradient.
+Note that as $h$ approaches zero, this approaches the actual value of the gradient; and has $h$ approaches infinity this value diverges.
+Figure 4 illustrates the impact of $h$ on the accuracy of the gradient: for small enough $h$, the approximation is accurate to several orders of magnitude.
+
+\begin{figure}[ht!]
+  \centering
+    \begin{tabular}{c | c}
+    $h$  & $\nabla_h[f](x)$ \\	\hline
+	0.01 & $-3.1704 \cdot 10^{-7}$ \\
+	0.1 & $-3.1704 \cdot 10^{-7}$ \\
+	1.0 & $-3.1696 \cdot 10^{-7}$ \\
+	10.0 & $-3.0923 \cdot 10^{-7}$ \\
+	100.0 & $-2.6198 \cdot 10^{-8}$ \\
+	1000.0 & $\approx 0$ \\
+    \end{tabular}
+  \caption{As $h$ increases, the accuracy of the central finite difference decreases. The actual gradient value is $-3.1704 \cdot 10^{-7}$.}
+\end{figure}
+
 \end{document}

### 2016-09-30 paper2/paper.tex
merge work

--- 2016-09-29 paper2/paper.tex
@@
-\documentclass[letterpaper, twocolumn,11pt]{article}
-\usepackage{epsfig,endnotes,amsmath,usenix}
+\documentclass[10pt]{paper}
+\usepackage{bm,amsmath,fullpage,graphicx,chngcntr}
 
-\title{\textbf{Regressions And Stuff}}
+\counterwithin{figure}{section}
 
-\author{
-\rm{Vahid Fazel-Rezai}\\
-\rm{\texttt{vfazel@mit.edu}}
-\and 
-\rm{Akshay Ravikumar}\\
-\rm{\texttt{akshayr@mit.edu}}
-}
-
-\date{May 6, 2016}
+\title{Regressions And Stuff}
 
 \begin{document}
-% \begin{titlepage}
-% \centering
-% Matei Zaharia\\
-% TR 10\\
-% May 6, 2016
-% \end{titlepage}
 
 \maketitle
-\section{Introduction}
 
-In this paper, we discuss various methods of calculating regression.
-First, we discuss the benfits of using Stochastic Gradient Descent versus Batch Gradient Descent.
-Next, we analyze the performance of these methods in performing a Least Squares Linear regression.
-Finally, we discuss blah blah
+\section{Function Basis}
+Consider a basic prediction problem where given an input, $x$, we wish to predict a value $y$ (e.g.
+predict income given GPA, or predict crime rate given temperature).
+We train on a sample of labeled data (a set of $(x,y)$ pairs) to find a function $f(x)$ that will give the expected $y$ for new inputs $x$.
 
-\section{Overview}
+One approach to this problem is to model $f$ as the linear combination of \emph{basis functions}.
+In other words, we manually choose functions $\phi_1(x), \dots, \phi_n(x)$ and algorithmically pick coefficients $c_1, \dots, c_n$, which we put together as
+\begin{align*}bm
+f(x) &= c_1 \phi_1(x) + \dots + c_n \phi_n(x).
+\end{align*}
+We can rewrite this with 
+\begin{align*}
+\bm{c} &= \begin{bmatrix} c_1 & \cdots & c_n \end{bmatrix} \\
+\bm{\phi}(x) &= \begin{bmatrix} \phi_1(x) & \cdots & \phi_n(x) \end{bmatrix}
+\end{align*}
+so that 
+\begin{align}
+f(x) &= \bm{c} \cdot \bm{\phi}(x). \label{eq:predictor}
+\end{align}
 
-\section{Gradient Descent}
-\subsection{Batch Gradient Descent}
-\subsection{Stochastic Gradient Descent}
-\subsection{Effect of Finite Gradient}
+The choice of the basis functions is somewhat arbitrary and often depends on the problem domain. Common choices are monomials ($\phi_i(x) = x^i$) and sinusoidals ($\phi_i(x) = \sin(i \pi x)$).
+
+As for the algorithm to pick the coefficient vector $\bm{c}$, we discuss three methods: \emph{linear regression}, \emph{ridge regression}, and \emph{LASSO} (least absolute shrinkage and selection operator).
+We devote one section to exploring each method in the next three sections.
+
+Before moving on, we develop common notation to set up the algorithmic problem.
+Suppose we are given the training data as column vectors $\bm{x}$ and $\bm{y}$ of length $m$ and have chosen $n$ basis functions given by $\bm{\phi}$.
+We let $X = \bm{\phi}(\bm{x})$ be the $m$ by $n$ matrix where the $i$-th row is $\bm{\phi}(x_i)$ and $x_i$ is the $i$-th element of $\bm{x}$.
+Thus, we can roughly frame our algorithm's goal as finding a column vector $\bm{c}$ of size $n$ such that
+\begin{align}
+\bm{y} &\approx X \bm{c}, \label{eq:problem}
+\end{align}
+where $X$ and $\bm{y}$ are taken as inputs.
+
+\section{Linear Regression}
+
+Linear regression is a regression method that simply aims to minimize the square of the magnitude of the difference of the two sides of (\ref{eq:problem}):
+\begin{align}
+\left| \bm{y} - X \bm{c} \right|^2. \label{eq:sse}
+\end{align}
+The expression in (\ref{eq:sse}) is known as the \emph{sum of squares error}, or SSE, as it is the sum of the squares of the differences in predicted versus real $y$ values.
+
+One way to find the optimal $\bm{c}$ is to expand the expression and set its partial derivative with respect to $\bm{c}$ equal to zero. Doing so yields
+\begin{align}
+\bm{c} &= \left( X^T X \right)^{-1} X^T \bm{y}. \label{eq:linearclosed}
+\end{align}
+We refer to (\ref{eq:linearclosed}) as the closed form for $\bm{c}$.
+
+Another way to find $\bm{c}$ is to perform gradient descent with SSE as the objective function of $\bm{c}$.
+
+We will illustrate and contrast these methods with an example, and then show the effects of choosing a different function basis.
+Our example is generated by taking $\bm{x}$ as numbers from 0 to 1 at intervals of $0.1$ and
+\begin{align*}
+\bm{y} &= \cos(\pi \bm{x}) + \cos(2 \pi \bm{x}) + \text{noise}.
+\end{align*}
+
+\subsection{Polynomials with Closed Form}
+
+We use $\bm{\phi}$ of size $n+1$ whose components are monomials of order 0 through $n$ as our funciton basis.
+Using this to compute $X$ and evaluate the closed form, we can obtain $\bm{c}$.
+We view the results as a plot of the training data points, the generating curve (without noise), and the predictor $f(x) = \bm{c} \cdot \bm{\phi}(x)$.
+This is shown in Figure \ref{fig:poly} for several values of $n$.
+
+\begin{figure}[ht!]
+  \centering
+  \label{fig:poly}
+  \includegraphics[width=0.45\textwidth]{../images/poly0.png}
+  \includegraphics[width=0.45\textwidth]{../images/poly1.png}
+  \includegraphics[width=0.45\textwidth]{../images/poly3.png}
+  \includegraphics[width=0.45\textwidth]{../images/poly10.png}
+  \caption{Training data (green), generating curve (red), and predicted curve (blue) for various $n$.}
+\end{figure}
+
+We can see that increasing the dimension of the function basis improves the training accuracy of our predictor.
+
+\subsection{Polynomials with Gradient Descent}
+
+Use batch gradient descent on the SSE function on some values of M. Describe your experience
+with initial guesses, step sizes and convergence thresholds. Compare with using SGD on the
+same data. Explain your results in terms of the properties of the function being optimized and
+the properties of the algorithms.
+
+\subsection{Cosines with Closed Form and Gradient Descent}
+
+We now try using the function basis consisting of $\cos(\pi x), \cos(2 \pi x), \dots, \cos(n \pi x)$.
+In this case the closed form and both versions of gradient descent all give the same coefficient vectors after converging.
+We plot in Figure \ref{fig:cos} comparisons for a couple values of $n$.
+
+\begin{figure}[ht!]
+  \centering
+  \label{fig:cos}
+  \includegraphics[width=0.45\textwidth]{../images/cos2.png}
+  \includegraphics[width=0.45\textwidth]{../images/cos8.png}
+  \caption{Training data (green), generating curve (red), and predicted curve (blue) for various $n$.}
+\end{figure}
+
+A notable observation is that even when the same basis functions were used as in the generating function ($\cos(\pi x)$ and $\cos(2\pi x)$ when $n=2$), we obtained a slightly different predicted curve.
+This is due to the added noise.
+
+Also, when we increase the dimension of the basis ($n=8$) the predicted curve becomes more complex.
+Although it is more accurate for the training data, it is more sensitive to the noise and so wouldn't generalize well to new test data.
+This behavior is called \emph{overfitting}, which results when the coefficients become too large or when too many features (basis functions) are used.
+We will see below that these issues can be addressed using ridge regression and LASSO, respectively.
+
+\section{Ridge Regression}
+
+Ridge regression is similar to linear regression in that its choice of $\bm{c}$ aims to minimize SSE, but in addition also aims to minimize the size of the coefficients themselves.
+Tangibly, this is achieved by choosing $\bm{c}$ that gives the minimal value of
+\begin{align}
+\left| \bm{y} - X \bm{c} \right|^2 + \lambda \left| \bm{c} \right|^2. \label{eq:ridgeobjective}
+\end{align}
+We can tune the sensitivity to coefficient sizes by adjusting the parameter $\lambda$.
+At the extreme, $\lambda = 0$ will have no consideration of coefficient size and will be the same as linear regression.
+
+Again, there are two ways to arrive at the desired $\bm{c}$ from here. First, we can expand (\ref{eq:ridgeobjective}) and set the partial derivative with respect to $\bm{c}$ equal to zero, which gives
+\begin{align*}
+\bm{c} &= \left( X^T X + \lambda I \right)^{-1} X^T \bm{y}. \label{eq:ridgeclosed}
+\end{align*}
+Second, we can perform gradient descent and converge to the optimal value of $\bm{c}$ numerically.
+
+In general, the advantage of the second term in the objective function is that resulting coefficients are smaller, thus empirically less sensitive to noise and more accurate on unseen data.
+
+\section{LASSO}
 

### 2016-09-30 code/P2/problem2.py

@@ -50,18 +50,18 @@ def add_poly_plot(M):
     def curve(x):
         return np.dot(weights, fn(x))
     plot_curve(curve, min(X), max(X), 3)
-    plt.title('Linear Regression (M=' + str(M) + ')' )
+    plt.title('Linear Regression (n=' + str(M) + ')' )
 
-def add_fn_plot(weights, fn):
+def add_fn_plot(weights, fn, width=2):
     def curve(x):
         return np.dot(weights, fn(x))
-    plot_curve(curve, min(X), max(X), 3)
+    plot_curve(curve, min(X), max(X), width)
 
 def make_plot():
     plt.plot(X, Y, 'o')
     plot_curve(generator, min(X), max(X), 1)
     plt.xlabel('x')
-    plt.xlabel('y')
+    plt.ylabel('y')
     plt.show()
 
 def sse_objective(fn):
@@ -84,23 +84,22 @@ def compare(fn, M, step, threshold):
 
     weights1 = regress(fn)
     weights2 = gradient_descent(sse_objective(fn), sse_gradient(fn), guess, step, threshold)
-    #weights3 = stochastic_gradient_descent(sse_objective(fn), sse_gradient(fn), guess, step, threshold)
+    # weights3 = stochastic_gradient_descent(sse_objective(fn), sse_gradient(fn), guess, step, threshold)
 
     print weights1
     print weights2
     # print weights3
 
     add_fn_plot(weights1, fn)
-    add_fn_plot(weights2, fn)
+    # add_fn_plot(weights2, fn, 2)
     # add_fn_plot(weights3, fn)
 
     make_plot()
 
 
 ##### 2.1
-# add_poly_plot(5)
+# add_poly_plot(10)
 # make_plot()
-
 
 ##### 2.2
 # see sse_objective and sse_gradient above
@@ -110,5 +109,6 @@ def compare(fn, M, step, threshold):
 # compare(poly_fn(M), M, .001, .000001)
 
 ##### 2.4
-# M = 5
-# compare(cos_fn(M), M, .001, .000001)
+M = 7
+plt.title('Linear Regression (n=' + str(M+1) + ')' )
+compare(cos_fn(M), M, .001, .000001)

### 2016-09-30 paper2/paper.tex

@@
 \documentclass[10pt]{paper}
-\usepackage{bm,amsmath,fullpage,graphicx,chngcntr}
-
+\usepackage{bm,amsmath,graphicx,chngcntr,subfigure}
+\usepackage[margin=0.5in]{geometry}
+\pagestyle{empty}
 \counterwithin{figure}{section}
-
+\pagenumbering{gobble}
 \title{Regressions And Stuff}
@@
 \maketitle
+\section{Introduction}
+
+In this paper, we discuss various methods of calculating regression.
+First, we discuss the benfits of using Stochastic Gradient Descent versus Batch Gradient Descent.
+Next, we analyze the performance of these methods in performing a Least Squares Linear regression.
+Finally, we discuss blah blah
+
+\section{Gradient Descent}
+
+\emph{Gradient descent} is an iterative optimization algorithm; in other words, it calculates the parameters $\mathbf{x}$ that minimize a given objective function $f(\mathbf{x})$.
+The algorithm repeatedly translates an initial \emph{guess} $\mathbf{x}_0$ in a direction proportional to the negative \emph{gradient} $\nabla f(x)$.
+
+In every step of the itreration, we update our guess as following:
+\begin{align*}
+\mathbf{x}_{n+1} = \mathbf{x} - \lambda \nabla f(\mathbf{x}_{n})
+\end{align*}
+where $\lambda$ is the \emph{step size} of the iteration. The algorithm terminates upon the following convergence condition: 
+\begin{align*}
+|f(\mathbf{x}_{n+1}) - f(\mathbf{x}_{n})| < \delta
+\end{align*}
+where $\delta$ is the convergence \emph{threshold}. Upon convergence, the algorithm returns a final guess of $\mathbf{x}_{\text{opt}} = \mathbf{x}_{n+1}$.
+
+In this section, we compare two methods of gradient descent: \emph{batch gradient descent}, which computes a gradient over all samples for each iteration, and \emph{stochastic gradient descent}, which instead uses pointwise gradients.
+
+To analyze the perforamnce of the different algorithms, we use the \emph{Gaussian function}:
+\begin{align*}
+f(x) = - \dfrac{1}{\sqrt{(2\pi)^n |\Sigma|}} \exp \left[-\dfrac{1}{2} (x - u)^T \Sigma^{-1} (x-u)\right]
+\end{align*}
+In addition, we use the \emph{quadratic bowl function}:
+\begin{align*}
+f(x) = \dfrac{1}{2} x^T A x - x^T b 
+\end{align*}
+Finally, we use the \emph{least squares error} function:
+\begin{align*}
+J(\theta) = |X\theta - y|^2
+\end{align*}
+
+\subsection{Batch Gradient Descent}
+
+In batch gradient descent, for every iteration we calculate the gradient across the entire sample set.
+In this section, we analyze the affects of step size ($\lambda$), threshold ($\delta$), and initial guess $\mathbf{x}_{0}$ on the performance and accuracy of the gradient descent algorithm.
+
+First, we note that step size has a significant impact on the convergence rate of the algorithm.
+We run gradient descent on an arbitrary quadratic bowl function.
+As we increase the value of $\lambda$ from $0.0001$ to $0.001$, the magnitude of the gradient decreases more quickly, and the algorithm converges faster.
+Note that in this figure, all of the gradient descents were run with a fixed size of 2000 iterations to illustrate the effects on the gradient over time.
+
+In addition, the step size and threshold control the accuracy of the gradient descent.
+As step size decreases, we expect a less accurate result because the gradient descent will likely converge at the boundary of the threshold, which a larger step size might lead to convergence well past the threshold.
+In addition, as the threshold decreases, we expect a more accurate result because the algorithm needs to run for longer before reaching a stable equilibrium.
+To verify these claims, we run gradient descent using a two-dimensional Gaussian with mean $(10, 10)$.
+As we vary $\lambda$ and $\delta$, we plot the least-squares error between the returned value and the actual optimum of $(10, 10)$ (see Figure 2).
+Indeed, the error decreases as $\lambda$ increases and $\delta$ decreases.
+
+Finally, we observe the affect of the initial case on the rate of convergence.
+When running gradient descent on a two-dimensional Gaussian with mean $(10, 10)$, one would expect that as the intial guess appraoches the optimal value, the convergence rate increases.
+Indeed, running the algorithm with $\lambda = 1000$ and $\delta = 10^{-11}$, we notice that the rate of convergence is much higher as the initial guess approaches the actual value (see Figure 3).
+
+\begin{figure}[ht!]
+  \centering
+  \begin{subfigure}
+    \centering
+      \includegraphics[width=0.3\textwidth]{../images/quadratic_steps_vs_mag}
+    % \caption{As the step size increases, the algorithm converges quicker.}
+  \end{subfigure}
+  \begin{subfigure}
+    \centering
+      \includegraphics[width=0.3\textwidth]{../images/lsq_error_gradient}
+    % \caption{As $\lambda$ increases and $\delta$ decreases, error decreases.}
+  \end{subfigure}
+  \begin{subfigure}
+    \centering
+      \includegraphics[width=0.3\textwidth]{../images/guess_vs_mags_2}
+    % \caption{As $\mathbf{x}_0$ approaches $(10, 10)$, the rate of convergence increases.}
+  \end{subfigure}
+\end{figure}
+
+\subsection{Stochastic Gradient Descent}
+
+Sometimes, it might be inefficient to compute a gradient over the entire dataset for every iteration, especially when the dataset is large.
+One way to address this is to use stochastic gradient descent, which uses a pointwise gradient instead.
+In this experiment, we use least-squares regression, which has a pointwise gradient of
+
+\begin{align*}
+J(\theta; x^{(i)}, y^{(i)}) = (x^{(i)}\theta_t - y^{(i)})^2
+\end{align*}
+On each iteration, we update our value as follows:
+\begin{align*}
+\theta_{t+1} = \theta_{t} - \nu_t \nabla_\theta J(\theta_t; x^{(i)}, y^{(i)})
+\end{align*}
+where $\nabla = (\tau_0 + t)^{-\kappa}$ is the \emph{learning rate}, and $\tau_0$ and $\kappa$ are constants.
+
+The benefit of stochastic gradient descent is that computations are quicker, and thus the algorithm converges faster, with respect to both time and the number of iterations.
+However, this comes with a decrease in performance, because each iteration uses less information.
+Refer to Figure 4 for the difference in the magnitude in the gradient between the two methods: note that stochastic gradient descent is much noisier, and in the end, does not converge to as accurate of a value as batch gradient descent does.
+
+\begin{figure}[ht!]
+  \center
+  \begin{subfigure}
+    \centering
+      \includegraphics[width=0.45\textwidth]{../images/stochastic_gradient_steps}
+    % \caption{Compare to batch gradient descent, the magnitude of the gradient during stochastic gradient descent shows significant noise.}
+  \end{subfigure}
+  \begin{subfigure}
+    \centering
+      \includegraphics[width=0.45\textwidth]{../images/batch_gradient_steps}
+    % \caption{The magnitude of the gradient during batch gradient descent is much smoother.}
+  \end{subfigure}
+\end{figure}
+
+\subsection{Effect of Finite Gradient}
+
+When the objective function doesn't have a closed form solution, one can approximate the gradient using a central finite difference. In other words, we have 
+\begin{align*}
+\nabla f(x) \approx \nabla_h[f](x) = \dfrac{f(x+h/2) - f(x-h/2)}{h}
+\end{align*}
+Therefore, we only need the objective function to approximate the gradient.
+Note that as $h$ approaches zero, this approaches the actual value of the gradient; and has $h$ approaches infinity this value diverges.
+Figures 4 and 5 illustrate the impact of $h$ on the accuracy of the gradient: even for relatively large $h$ (especially for the quadratic bowl), the approximation is accurate to several orders of magnitude.
+
+\begin{figure}[ht!]
+  \centering
+  \begin{subfigure}
+    \centering
+      \begin{tabular}{c | c}
+      $h$  & $\nabla_h[f](x)$ \\  \hline
+    0.01 & $-3.1704 \cdot 10^{-7}$ \\
+    0.1 & $-3.1704 \cdot 10^{-7}$ \\
+    1.0 & $-3.1696 \cdot 10^{-7}$ \\
+    10.0 & $-3.0923 \cdot 10^{-7}$ \\
+    100.0 & $-2.6198 \cdot 10^{-8}$ \\
+    1000.0 & $\approx 0$ \\
+      \end{tabular}
+    % \caption{As $h$ increases, the accuracy of the central finite difference for this Gaussian decreases. The actual gradient value is $-3.1704 \cdot 10^{-7}$.}
+  \end{subfigure}
+  \qquad
+  \begin{subfigure}
+    \centering
+      \begin{tabular}{c | c}
+      $h$  & $\nabla_h[f](x)$ \\  \hline
+    $10^{9}$ & $-280.000000$ \\
+    $10^{11}$ & $-279.999991$ \\
+    $10^{13}$ & $-280.000944$ \\
+    $10^{15}$ & $-280.349077$ \\
+    $10^{17}$ & $-299.759591$ \\
+    $10^{19}$ & $\approx 0$ \\
+      \end{tabular}
+    % \caption{As $h$ increases, the accuracy of the central finite difference for this quadratic bowl decreases very slowly. The actual gradient value is $-280$.}
+  \end{subfigure}
+\end{figure}
 
@@
   \label{fig:poly}
-  \includegraphics[width=0.45\textwidth]{../images/poly0.png}
-  \includegraphics[width=0.45\textwidth]{../images/poly1.png}
-  \includegraphics[width=0.45\textwidth]{../images/poly3.png}
-  \includegraphics[width=0.45\textwidth]{../images/poly10.png}
-  \caption{Training data (green), generating curve (red), and predicted curve (blue) for various $n$.}
+  \includegraphics[width=0.24\textwidth]{../images/poly0.png}
+  \includegraphics[width=0.24\textwidth]{../images/poly1.png}
+  \includegraphics[width=0.24\textwidth]{../images/poly3.png}
+  \includegraphics[width=0.24\textwidth]{../images/poly10.png}
+  \caption{Training data (green), generating curve (red), and predicted curve (blue) for $n = 0, 1, 3, 10$.}
 \end{figure}
@@
   \label{fig:cos}
-  \includegraphics[width=0.45\textwidth]{../images/cos2.png}
-  \includegraphics[width=0.45\textwidth]{../images/cos8.png}
-  \caption{Training data (green), generating curve (red), and predicted curve (blue) for various $n$.}
+  \includegraphics[width=0.3\textwidth]{../images/cos2.png}
+  \includegraphics[width=0.3\textwidth]{../images/cos5.png}
+  \includegraphics[width=0.3\textwidth]{../images/cos8.png}
+  \caption{Training data (green), generating curve (red), and predicted curve (blue) for $n=2, 5, 8$.}
 \end{figure}
@@
 
+\subsection{Continuation of Simple Example}
+
+In this subsection we apply ridge regression to the same example as the previous section, in which linear regression showed some overfitting.
+Varying $\lambda$ and $n$ gives us the plots in Figure \ref{fig:ridge}.
+
+\begin{figure}[ht!]
+  \centering
+  \label{fig:ridge}
+  \includegraphics[width=0.24\textwidth]{../images/ridge3low.png}
+  \includegraphics[width=0.24\textwidth]{../images/ridge3high.png}
+  \includegraphics[width=0.24\textwidth]{../images/ridge10low.png}
+  \includegraphics[width=0.24\textwidth]{../images/ridge10high.png}
+  \caption{Training data (green), generating curve (red), and predicted curve (blue) for $\lambda=0.0001, 0.1$ and $n = 3, 10$.}
+\end{figure}
+
+We can see that a good value of $\lambda$ gives a decent fit, while $\lambda$ too large results in \emph{underfitting}, where the prediction is not sensitive enough to the training data.
+
+\subsection{Training and Validation}
+
+As we just illustrated, choosing a good value of $\lambda$ is crucial for generalizing well.
+We now present a process for finding a good value for $\lambda$ (as well as $n$).
+Such a process is called \emph{model selection}.
+
+We divide the given labeled data into two groups, one called the training set and one called the validation set.
+We vary $n$ and $\lambda$, and for each combination we train as above on the training set and compute SSE on the validation set.
+Finally, we select the combination of parameters that resulted in the smallest validation SSE.
+
+To illustrate in an example, we have three sets of data: set A, set B, and a validation set.
+We perform the above process twice, once training on A and testing on B, and once vice versa.
+In both cases we try a variety of values of $\lambda$ and $n$ then evaluate the SSE on the validation set.
+The results are given in Figure \ref{fig:validationtable}
+\begin{figure}[ht!]
+  \centering
+  \label{fig:validationtable}
+  \begin{tabular}{c|cccccc}
+    & 1e-05 & 0.0001 & 0.001 & 0.01 & 0.1 \\ \hline
+  0 & 62.00 & 62.00 & 62.00 & 62.01 & 62.06 \\
+  1 & 2.90 & 2.90 & 2.90 & 2.90 & 2.94 \\
+  2 & 2.35 & \textbf{2.35} & 2.35 & 2.36 & 2.43 \\
+  3 & 2.37 & 2.37 & 2.37 & 2.39 & 2.55 \\
+  4 & 3.58 & 3.58 & 3.58 & 3.59 & 3.67 \\
+  5 & 4.14 & 4.14 & 4.14 & 4.12 & 3.97 \\
+  6 & 3.98 & 3.98 & 3.98 & 3.96 & 3.89 \\
+  7 & 4.07 & 4.07 & 4.06 & 4.03 & 3.89 \\
+  \end{tabular}
+  \qquad
+  \begin{tabular}{c|ccccc}
+    & 0 & 0.3 & 0.6 & 0.9 & 1.2 \\ \hline
+  0 & 72.44 & 70.95 & 69.63 & 68.47 & 67.44 \\
+  1 & 35.19 & 34.09 & 33.14 & 32.34 & 31.67 \\
+  2 & 38.53 & 38.65 & 38.76 & 38.87 & 38.96 \\
+  3 & 30.30 & 28.46 & \textbf{28.09} & 28.47 & 29.21 \\
+  4 & 130.20 & 67.54 & 51.63 & 46.17 & 44.35 \\
+  5 & 625.43 & 368.40 & 287.26 & 242.74 & 212.14 \\
+  6 & 680.05 & 533.60 & 481.26 & 440.20 & 406.68 \\
+  7 & 492.85 & 360.39 & 448.42 & 544.91 & 634.85 \\
+  \end{tabular}
+  \caption{SSE values on validation set after training with various $n$ (rows) and $\lambda$ (columns). Left: training on set A. Right: training on set B. Best values bolded.}
+\end{figure}
+The SSE for the first case with $n = 2$ and $\lambda = 0.0001$, tested on set B, is $25.752$.
+In the second case, tested on set A with $n = 3$ and $\lambda = 0.6$, the SSE is $40.087483697$.
+Testing with parameters $n=4$ and $\lambda = 0.1$ for both cases, which the table would predict to be suboptimal, indeed gives bigger test SSEs of 31.1388659319 and 64.9392925156, respectively.
+
+We can see in this example that the choice of training set matters. Also, we note that the best values for the parameters $n$ and $\lambda$ depends heavily on the given data.
+
 \section{LASSO}

### 2016-09-30 code/P2/problem2.py
committing

@@ -108,7 +108,21 @@ def compare(fn, M, step, threshold):
 # M = 3
 # compare(poly_fn(M), M, .001, .000001)
 
-##### 2.4
-M = 7
+#### 2.4
+M = 4
 plt.title('Linear Regression (n=' + str(M+1) + ')' )
 compare(cos_fn(M), M, .001, .000001)
+
+
+def ridge(fn, lam):
+    Z = np.array([fn(x) for x in X])
+    return np.dot(np.dot(np.linalg.inv(np.add(np.dot(np.transpose(Z), Z), lam * np.identity(Z.shape[1]))), np.transpose(Z)), Y)
+
+def simple_ridge(M, lam):
+	fn = poly_fn(M)
+	weights = ridge(fn, lam)
+	obj = sse_objective(fn)
+	return obj(weights)
+
+
+# print simple_ridge(3, 0.1)

### 2016-09-30 code/P3/problem3.py

@@ -1,0 +1,62 @@
+import math
+import numpy as np
+import matplotlib.pyplot as plt
+import pylab as pl
+
+def getData(name):
+    data = pl.loadtxt(name)
+    # Returns column matrices
+    X = np.array(data[0:1])[0]
+    Y = np.array(data[1:2])[0]
+    return X, Y
+
+def regressAData():
+    return getData('P3/regressA_train.txt')
+
+def regressBData():
+    return getData('P3/regressB_train.txt')
+
+def validateData():
+    return getData('P3/regress_validate.txt')
+
+
+# returns function that evaluates as a polynomial with given weights
+def poly_fn(M):
+    def poly(x):
+        return np.array([(x ** i) for i in range(M+1)])
+    return poly
+
+def sse(X, Y, fn, weights):
+    return np.linalg.norm(Y - np.dot(weights, fn(X))) ** 2
+
+def ridge(M, lam, X, Y):
+    Xv, Yv = validateData()
+    fn = poly_fn(M)
+    Z = np.array([fn(x) for x in X])
+    weights = np.dot(np.dot(np.linalg.inv(np.add(np.dot(np.transpose(Z), Z), lam * np.identity(Z.shape[1]))), np.transpose(Z)), Y)
+    return sse(Xv, Yv, fn, weights)
+
+# lams1 = [0, 0.3, 0.6, 0.9, 1.2]
+# lams2 = [0.00001, 0.0001, 0.001, 0.01, 0.1]
+# X, Y = regressAData()
+# lams = lams2
+# out = '   '
+# for lam in lams:
+#     out += ' & ' + str(lam)
+# out += ' \\\\ \\hline\n'
+# for M in [0, 1, 2, 3, 4, 5, 6, 7]:
+#     out += '  ' + str(M)
+#     for lam in lams:
+#         cur = ridge(M, lam, X, Y)
+#         out += ' & ' + '%.2f'%(cur)
+#     out += ' \\\\\n'
+# print out
+
+M = 4
+lam = 10
+X, Y = regressAData()
+Xv, Yv = regressBData()
+fn = poly_fn(M)
+Z = np.array([fn(x) for x in X])
+weights = np.dot(np.dot(np.linalg.inv(np.add(np.dot(np.transpose(Z), Z), lam * np.identity(Z.shape[1]))), np.transpose(Z)), Y)
+print sse(Xv, Yv, fn, weights)

### 2016-09-30 paper2/paper.tex
2.3

@@
     \centering
+      \includegraphics[width=0.3\textwidth]{../images/guess_vs_mags_2}
+    % \caption{As $\mathbf{x}_0$ approaches $(10, 10)$, the rate of convergence increases.}
+  \end{subfigure}
+  \begin{subfigure}
+    \centering
       \includegraphics[width=0.3\textwidth]{../images/lsq_error_gradient}
@@
   \end{subfigure}
-  \begin{subfigure}
-    \centering
-      \includegraphics[width=0.3\textwidth]{../images/guess_vs_mags_2}
-    % \caption{As $\mathbf{x}_0$ approaches $(10, 10)$, the rate of convergence increases.}
-  \end{subfigure}
+  \caption{The effects of $\lambda$, $\mathbf{x}_0$, and $\lambda$ on gradient descent.}
 \end{figure}
@@
     \centering
-      \includegraphics[width=0.45\textwidth]{../images/batch_gradient_steps}
+      \includegraphics[width=0.45\textwidth]{../images/batch_gradient_steps_2}
     % \caption{The magnitude of the gradient during batch gradient descent is much smoother.}
   \end{subfigure}
+  \caption{Stochastic gradient descent (left) is much noisier than batch gradient descent (right).}
 \end{figure}
@@
 
-Use batch gradient descent on the SSE function on some values of M. Describe your experience
-with initial guesses, step sizes and convergence thresholds. Compare with using SGD on the
-same data. Explain your results in terms of the properties of the function being optimized and
-the properties of the algorithms.
+In this section, we compare the performance of batch and stochastic gradient descent on linear regression.
+See Figure \ref{fig:gd_m} for the performance of the two gradient descent algorithms on a polynomial function.
+It appears that, even for high values of $M$ (the degree of the polynomial), the results of either gradient descent are very inaccurate -- in fact, they essentially look quadratic.
+This is because the values of $x^k$ have extremely small values as $k$ increases for $|x| < 1$.
+Therefore, the gradient has a very small effect for higher degrees, while smaller degrees (i.e.
+$k = 0, 1, 2$) have a much larger impact.
+
+To counteract this effect, the learning rate for the stochastic gradient descent needs to be as low as possible for.
+In this case, we chose $\nu_t = (r + 0.1)^{-0.51}$; with parameters any larger, the result has very little correlation with the actual function.
+This is likely because a large learning rate would cause smaller-degree terms to have an relatively large weight during the first few iterations, which would overtake the higher-degree terms.
+Having a smaller, slower-growing learning rate lets the higher-degree terms affect the result for more iterations, and therefore the curve can fit a larger number of points.
+
+Finally, initial guess and threshold have very little impact on the accuracy of the function; they simply have an affect on performance.
+
+\begin{figure}[ht!]
+  \centering
+  \label{fig:gd_m}
+  \includegraphics[width=0.24\textwidth]{../images/all_m_3}
+  \includegraphics[width=0.24\textwidth]{../images/all_m_5}
+  \includegraphics[width=0.24\textwidth]{../images/all_m_7}
+  \includegraphics[width=0.24\textwidth]{../images/all_m_10}
+  \caption{The results for GD and SGD for $M = 3, 5, 7, 10$ (in that order).}
+\end{figure}
 

### 2016-09-30 paper2/paper.tex
paper 2 proofread

@@
 \pagenumbering{gobble}
-\title{Regressions And Stuff}
+\title{6.857 Homework 1: Gradient Descent and Regressions}
 
@@
 
-In this paper, we discuss various methods of calculating regression.
-First, we discuss the benfits of using Stochastic Gradient Descent versus Batch Gradient Descent.
-Next, we analyze the performance of these methods in performing a Least Squares Linear regression.
-Finally, we discuss blah blah
+In this paper, we discuss various methods of performing regressions.
+First, we present the stochastic gradient descent and batch gradient descent algorithms, and discuss the benefits and drawbacks of each.
+Next, we use gradient descent to perform a linear regression on functions with polynomial and sinusoidal bases.
+We then improve upon this model with ridge regression, which uses an extra weighting term to prevent overfitting.
+Finally, we present LASSO estimation, which optimizes this model for the case of sparse weights.
 
@@
 
-\emph{Gradient descent} is an iterative optimization algorithm; in other words, it calculates the parameters $\mathbf{x}$ that minimize a given objective function $f(\mathbf{x})$.
+We start by introducing \emph{gradient descent} iterative optimization algorithm, which calculates the parameters $\mathbf{x}$ that minimize a given objective function $f(\mathbf{x})$.
 The algorithm repeatedly translates an initial \emph{guess} $\mathbf{x}_0$ in a direction proportional to the negative \emph{gradient} $\nabla f(x)$.
+By moving in the direction of the negative gradient, the guess tends towards a local minimum.
 
-In every step of the itreration, we update our guess as following:
+In every step of the iteration, we update our guess as follows:
 \begin{align*}
@@
 
-To analyze the perforamnce of the different algorithms, we use the \emph{Gaussian function}:
+To analyze the perforamnce of the different algorithms, we use the following functions:
+
+\begin{enumerate}
+\item \emph{Gaussian} function:
 \begin{align*}
@@
 \end{align*}
-In addition, we use the \emph{quadratic bowl function}:
+\item \emph{Quadratic bowl} function:
 \begin{align*}
@@
 \end{align*}
-Finally, we use the \emph{least squares error} function:
+\item \emph{Least squares error} function:
 \begin{align*}
@@
 \end{align*}
+\end{enumerate}
 
@@
 
-In batch gradient descent, for every iteration we calculate the gradient across the entire sample set.
-In this section, we analyze the affects of step size ($\lambda$), threshold ($\delta$), and initial guess $\mathbf{x}_{0}$ on the performance and accuracy of the gradient descent algorithm.
+In batch gradient descent, we calculate the gradient across the entire sample set for each iteration.
+In this section, we analyze the affects of step size ($\lambda$), threshold ($\delta$), and initial guess $\mathbf{x}_{0}$ on the performance and accuracy of batch gradient descent.
 
 First, we note that step size has a significant impact on the convergence rate of the algorithm.
-We run gradient descent on an arbitrary quadratic bowl function.
-As we increase the value of $\lambda$ from $0.0001$ to $0.001$, the magnitude of the gradient decreases more quickly, and the algorithm converges faster.
-Note that in this figure, all of the gradient descents were run with a fixed size of 2000 iterations to illustrate the effects on the gradient over time.
+To test this, we run gradient descent on an arbitrary quadratic bowl function.
+As we increase the value of $\lambda$ from $0.0001$ to $0.001$, the magnitude of the gradient decreases more quickly, and the algorithm converges faster (see Figure \ref{fig:gd}, left).
+Note that in the figure, all of the gradient descents were run with a fixed size of 2000 iterations.
 
 In addition, the step size and threshold control the accuracy of the gradient descent.
-As step size decreases, we expect a less accurate result because the gradient descent will likely converge at the boundary of the threshold, which a larger step size might lead to convergence well past the threshold.
+As step size increases, we expect a more accurate result because the gradient descent will likely converge well within the threshold, while a smaller step size might lead to convergence at the boundary of the threshold.
 In addition, as the threshold decreases, we expect a more accurate result because the algorithm needs to run for longer before reaching a stable equilibrium.
-To verify these claims, we run gradient descent using a two-dimensional Gaussian with mean $(10, 10)$.
-As we vary $\lambda$ and $\delta$, we plot the least-squares error between the returned value and the actual optimum of $(10, 10)$ (see Figure 2).
+To verify this, we run gradient descent using a two-dimensional Gaussian with mean $(10, 10)$.
+As we vary $\lambda$ and $\delta$, we plot the least-squares error between the returned value and the actual optimum of $(10, 10)$ (see Figure \ref{fig:gd}, middle).
 Indeed, the error decreases as $\lambda$ increases and $\delta$ decreases.
 
-Finally, we observe the affect of the initial case on the rate of convergence.
-When running gradient descent on a two-dimensional Gaussian with mean $(10, 10)$, one would expect that as the intial guess appraoches the optimal value, the convergence rate increases.
-Indeed, running the algorithm with $\lambda = 1000$ and $\delta = 10^{-11}$, we notice that the rate of convergence is much higher as the initial guess approaches the actual value (see Figure 3).
+Finally, we observe the affect of the initial guess on the rate of convergence.
+When running gradient descent on a two-dimensional Gaussian with mean $(10, 10)$, one would expect that as the intial guess approaches the optimal value, the convergence rate increases.
+Running the algorithm with $\lambda = 1000$ and $\delta = 10^{-11}$, we notice that the rate of convergence is much higher as the initial guess approaches the optimal value (see Figure \ref{fig:gd}, right).
 
 \begin{figure}[ht!]
+  \label{fig:gd}
   \centering
@@
     \centering
+      \includegraphics[width=0.3\textwidth]{../images/lsq_error_gradient}
+    % \caption{As $\lambda$ increases and $\delta$ decreases, error decreases.}
+  \end{subfigure}
+   \begin{subfigure}
+    \centering
       \includegraphics[width=0.3\textwidth]{../images/guess_vs_mags_2}
@@
   \end{subfigure}
-  \begin{subfigure}
-    \centering
-      \includegraphics[width=0.3\textwidth]{../images/lsq_error_gradient}
-    % \caption{As $\lambda$ increases and $\delta$ decreases, error decreases.}
-  \end{subfigure}
-  \caption{The effects of $\lambda$, $\mathbf{x}_0$, and $\lambda$ on gradient descent.}
+  \caption{The effects of $\lambda$ (left), $\mathbf{x}_0$ (middle), and $\delta$ (right) on gradient descent.}
 \end{figure}
@@
 
-Sometimes, it might be inefficient to compute a gradient over the entire dataset for every iteration, especially when the dataset is large.
+With large datasets, it might be inefficient to compute a gradient over the entire dataset for each iteration.
 One way to address this is to use stochastic gradient descent, which uses a pointwise gradient instead.
 In this experiment, we use least-squares regression, which has a pointwise gradient of
-
 \begin{align*}
@@
 
-The benefit of stochastic gradient descent is that computations are quicker, and thus the algorithm converges faster, with respect to both time and the number of iterations.
-However, this comes with a decrease in performance, because each iteration uses less information.
-Refer to Figure 4 for the difference in the magnitude in the gradient between the two methods: note that stochastic gradient descent is much noisier, and in the end, does not converge to as accurate of a value as batch gradient descent does.
+The benefit of stochastic gradient descent is that computations are quicker, which means the algorithm converges in fewer iterations.
+However, because each iteration utilizes less information, this comes with a decrease in performance.
+Refer to Figure \ref{fig:gd_steps} for the difference in convergence rate between the two methods: note that stochastic gradient descent is much noisier, and in the end, does not converge to as accurate of a value as does batch gradient descent.
 
@@
   \center
+  \label{fig:gd_steps}
   \begin{subfigure}
@@
   \end{subfigure}
-  \caption{Stochastic gradient descent (left) is much noisier than batch gradient descent (right).}
+  \caption{Magnitude of the gradient vs. number of iterations. Stochastic gradient descent (left) is much noisier than batch gradient descent (right).}
 \end{figure}
@@
 
-When the objective function doesn't have a closed form solution, one can approximate the gradient using a central finite difference. In other words, we have 
+When the objective function doesn't have a closed form solution, one can approximate the gradient using a \emph{central finite difference}. In other words, we have 
 \begin{align*}
@@
 Therefore, we only need the objective function to approximate the gradient.
-Note that as $h$ approaches zero, this approaches the actual value of the gradient; and has $h$ approaches infinity this value diverges.
-Figures 4 and 5 illustrate the impact of $h$ on the accuracy of the gradient: even for relatively large $h$ (especially for the quadratic bowl), the approximation is accurate to several orders of magnitude.
+Note that as $h$ approaches zero, this approaches the actual value of the gradient; and as $h$ increaes, the value diverges.
+Figure \ref{fig:cfd}illustrates the impact of $h$ on the accuracy of the gradient: even for large $h$ (especially for the quadratic bowl), the approximation is accurate to several orders of magnitude.
 
@@
   \centering
+  \label{fig:cfd}
   \begin{subfigure}
@@
   \end{subfigure}
+  \caption{The change in central finite difference as the value of $h$ increases for Gaussian (left) and quadratic bowl (right). The real values are $-3.1704 \cdot 10^{-7}$ and $-280$, respectively.}
 \end{figure}
@@
 
-The choice of the basis functions is somewhat arbitrary and often depends on the problem domain. Common choices are monomials ($\phi_i(x) = x^i$) and sinusoidals ($\phi_i(x) = \sin(i \pi x)$).
+The choice of the basis functions is somewhat arbitrary and often depend on the problem domain. Common choices are monomials ($\phi_i(x) = x^i$) and sinusoidals ($\phi_i(x) = \sin(i \pi x)$).
 
 As for the algorithm to pick the coefficient vector $\bm{c}$, we discuss three methods: \emph{linear regression}, \emph{ridge regression}, and \emph{LASSO} (least absolute shrinkage and selection operator).
-We devote one section to exploring each method in the next three sections.
+We explore each method over the next three sections.
 
@@
 This is likely because a large learning rate would cause smaller-degree terms to have an relatively large weight during the first few iterations, which would overtake the higher-degree terms.
-Having a smaller, slower-growing learning rate lets the higher-degree terms affect the result for more iterations, and therefore the curve can fit a larger number of points.
+Having a smaller, slower-growing learning rate lets the higher-degree terms affect the result for more iterations, and therefore the result more closely fits the input data.
 
-Finally, initial guess and threshold have very little impact on the accuracy of the function; they simply have an affect on performance.
+Finally, initial guess and threshold have very little impact on the accuracy of the gradient descent; they simply have an affect on performance.
 

### 2016-09-30 paper2/paper.tex
small stuff
done

@@
 \documentclass[10pt]{paper}
-\usepackage{bm,amsmath,graphicx,chngcntr,subfigure}
-\usepackage[margin=0.5in]{geometry}
+\usepackage{bm,amsmath,graphicx,chngcntr,subfigure,float}
+\usepackage[margin=0.43in]{geometry}
 \pagestyle{empty}
@@
 \pagenumbering{gobble}
-\title{6.857 Homework 1: Gradient Descent and Regressions}
+\title{6.867 Homework 1: Gradient Descent and Regression Methods}
 
@@
 \maketitle
+\vspace{-50px}
+
 \section{Introduction}
@@
 First, we present the stochastic gradient descent and batch gradient descent algorithms, and discuss the benefits and drawbacks of each.
-Next, we use gradient descent to perform a linear regression on functions with polynomial and sinusoidal bases.
-We then improve upon this model with ridge regression, which uses an extra weighting term to prevent overfitting.
-Finally, we present LASSO estimation, which optimizes this model for the case of sparse weights.
+Next, we use gradient descent to perform linear regressions.
+We then improve upon this model with ridge regression, which uses an extra regularization term to prevent overfitting.
+Finally, we present LASSO estimation, which optimizes this model to encourage sparse coefficient vectors.
 
@@
   \end{subfigure}
-  \caption{The effects of $\lambda$ (left), $\mathbf{x}_0$ (middle), and $\delta$ (right) on gradient descent.}
+  \caption{The effects of $\lambda$ (left), $\delta$ (middle), and $\mathbf{x}_0$ (right) on gradient descent.}
 \end{figure}
@@
 \begin{align*}
-\theta_{t+1} = \theta_{t} - \nu_t \nabla_\theta J(\theta_t; x^{(i)}, y^{(i)})
+\theta_{t+1} = \theta_{t} - \eta_t \nabla_\theta J(\theta_t; x^{(i)}, y^{(i)})
 \end{align*}
@@
 Note that as $h$ approaches zero, this approaches the actual value of the gradient; and as $h$ increaes, the value diverges.
-Figure \ref{fig:cfd}illustrates the impact of $h$ on the accuracy of the gradient: even for large $h$ (especially for the quadratic bowl), the approximation is accurate to several orders of magnitude.
+Figure \ref{fig:cfd} illustrates the impact of $h$ on the accuracy of the gradient: even for large $h$ (especially for the quadratic bowl), the approximation is accurate to several orders of magnitude.
 
@@
 In other words, we manually choose functions $\phi_1(x), \dots, \phi_n(x)$ and algorithmically pick coefficients $c_1, \dots, c_n$, which we put together as
-\begin{align*}bm
+\begin{align*}
 f(x) &= c_1 \phi_1(x) + \dots + c_n \phi_n(x).
@@
 To counteract this effect, the learning rate for the stochastic gradient descent needs to be as low as possible for.
-In this case, we chose $\nu_t = (r + 0.1)^{-0.51}$; with parameters any larger, the result has very little correlation with the actual function.
+In this case, we chose $\eta_t = (r + 0.1)^{-0.51}$; with parameters any larger, the result has very little correlation with the actual function.
 This is likely because a large learning rate would cause smaller-degree terms to have an relatively large weight during the first few iterations, which would overtake the higher-degree terms.
@@
   4 & 3.58 & 3.58 & 3.58 & 3.59 & 3.67 \\
-  5 & 4.14 & 4.14 & 4.14 & 4.12 & 3.97 \\
-  6 & 3.98 & 3.98 & 3.98 & 3.96 & 3.89 \\
-  7 & 4.07 & 4.07 & 4.06 & 4.03 & 3.89 \\
   \end{tabular}
@@
   4 & 130.20 & 67.54 & 51.63 & 46.17 & 44.35 \\
-  5 & 625.43 & 368.40 & 287.26 & 242.74 & 212.14 \\
-  6 & 680.05 & 533.60 & 481.26 & 440.20 & 406.68 \\
-  7 & 492.85 & 360.39 & 448.42 & 544.91 & 634.85 \\
   \end{tabular}
@@
 
+LASSO extends linear regression as well, but in a slightly different way.
+The balance between minimizing error and minimizing coefficient size is achieved by finding the optimal $\bm{c}$ for the objective function
+\begin{align}
+|\bm{y} - X \bm{c}|^2 + \lambda \sum_{i=1}^{n} |c_i|.
+\end{align}
+In other words, we use the sum of the first order magnitudes of the coefficients rather than second order.
+
+This gives a similar benefit as ridge regression in that it can be tuned to generalize, avoiding overfitting and underfitting.
+However, it additionally has a subtle but important advantage: LASSO encourages coefficients to be exactly zero, resulting in sparse coefficient vectors.
+This can be useful when the basis dimension is large by simplifying and speeding up computations (as all zero coefficients can be ignored) and also allows for the strategy of throwing all possible basis functions at the regression and letting LASSO find which functions to use.
+
+To intuitively see why LASSO encourages exactly zero coefficients, roughly consider applying gradient descent.
+If at a certain point the gradient in one variable is larger than that of another, it would make the objective function (specifically sum of absolute values) smaller to transfer the combined sum of the two variables all to only one of the variables, making the other one zero.
+
+\subsection{Sparsity in LASSO versus Ridge Regression}
+
+We illustrate the sparsity property of LASSO in an example where all the data points were generated using a sparse set of coefficients on a given set of basis functions plus some noise.
+The data points are divided into training, validation, and test sets.
+We train prediction curves with the same basis functions using linear regression (when $\lambda = 0$), ridge regression (parameters chosen using validation), and LASSO (parameters chosen using validation).
+
+Inspecting the coefficient vectors numerically, we find that linear and ridge regression have all nonzero coefficients, but LASSO gives nonzero coefficients exactly for the same functions as the true generating coefficients.
+Indeed, in this case LASSO correctly identified the correct basis functions.
+This is the ``selection operator'' that makes up the last two letters of LASSO.
+
+If we plot all the data points, prediction curves, and generating (true) curve in Figure \ref{fig:everything}, we see that all regressions give a relatively good prediction of the data.
+All the methods have generally find coefficients that give a close approximation of the training data, and in this case don't overfit.
+Depending on the application, sparsity may or may not be desired, and the method for solving the prediction problem can be appropriately chosen with consideration of each method's unique properties.
+
+\begin{figure}[H]
+  \centering
+  \label{fig:everything}
+  \includegraphics[width=0.5\textwidth]{../images/everything.png}
+  \caption{Comparison of various prediction curves plotted with data points.}
+\end{figure}
+
 \end{document}

### 2016-09-30 code/P4/lassoTestData.m

@@ -1,6 +1,0 @@
-function [X,Y] = lassoTestData()
-
-data = importdata('lasso_test.txt');
-
-X = data(1,:);
-Y = data(2,:);

### 2016-09-30 code/P4/lassoTrainData.m

@@ -1,6 +1,0 @@
-function [X,Y] = lassoTrainData()
-
-data = importdata('lasso_train.txt');
-
-X = data(1,:);
-Y = data(2,:);

### 2016-09-30 code/P4/lassoValData.m

@@ -1,6 +1,0 @@
-function [X,Y] = lassoValData()
-
-data = importdata('lasso_validate.txt');
-
-X = data(1,:);
-Y = data(2,:);

### 2016-09-30 code/P4/problem4.py

@@ -1,0 +1,95 @@
+import math
+import numpy as np
+import matplotlib.pyplot as plt
+import pylab as pl
+import pdb
+import random
+from sklearn import linear_model
+
+# X is an array of N data points (one dimensional for now), that is, NX1
+# Y is a Nx1 column vector of data values
+
+def getData(name):
+    data = pl.loadtxt(name)
+    # Returns column matrices
+    X = np.array(data[0:1])[0]
+    Y = np.array(data[1:2])[0]
+    return X, Y
+
+def lassoTrainData():
+    return getData('P4/lasso_train.txt')
+
+def lassoValData():
+    return getData('P4/lasso_validate.txt')
+
+def lassoTestData():
+    return getData('P4/lasso_test.txt')
+
+def getTrueWeights():
+    data = pl.loadtxt('P4/lasso_true_w.txt')
+    return np.array(data)
+
+Xtrain, Ytrain = lassoTrainData()
+Xval, Yval = lassoValData()
+Xtest, Ytest = lassoTestData()
+xmin = min(min(Xtrain), min(Xval), min(Xtest))
+xmax = max(max(Xtrain), max(Xval), max(Xtest))
+
+
+def phi(x):
+    out = [x]
+    for i in range(12):
+        out.append(np.sin((i+1) * 0.4 * math.pi * x))
+    return np.array(out)
+
+def plot_curve(weights, label, width=1):
+    x = [i/100.0 for i in range(100 * int(math.floor(xmin)), 100 * int(math.ceil(xmax)))]
+    y = [np.dot(weights, phi(t)) for t in x]
+    plt.plot(x, y, lw=width, label=label)
+
+def plot_curve_fn(fn, label, width=1):
+    x = [i/100.0 for i in range(100 * int(math.floor(xmin)), 100 * int(math.ceil(xmax)))]
+    y = [fn(phi(t)) for t in x]
+    plt.plot(x, y, lw=width, label=label)
+
+def regress(X, Y):
+    Z = np.array([phi(x) for x in X])
+    return np.dot(np.dot(np.linalg.inv(np.dot(np.transpose(Z), Z)), np.transpose(Z)), Y);
+
+def ridge(X, Y, lam):
+    Z = np.array([phi(x) for x in X])
[…]
